\emph{icepack}: a new glacier flow modeling package in Python, version 1.0

### 2021-04-03 reviews/response-2.tex

The fundamental physics are the first-order model obtained by asymptotic expansion of the Stokes equations in the aspect ratio, also known as the Blatter-Pattyn equations.
What I had imagined is that using only vertical basis functions up to degree 2 essentially defines its own semi-discrete physics model, similar to two- or three-layer ocean models.
You can view these as very coarse discretizations of the primitive equations, or you can view them as simplified models in their own right.
But I made this naming choice before there were many other collaborators on the project.
The ensuing confusion has shown that this was a bad choice of terminology that we'll change in a future version.
This model can be used for both regions at continental scales -- it can capture both plug and shear flow.

\item \textcolor{blue}{Section 2.2.1: in my opinion, the authors do not provide sufficient justification for the penalty term,
equation (7). They describe this as something that is added to smooth over artifacts - this would be
needed based on the discretization, which there is little discussion of. The authors should state what
order finite elements they are using - I presume they are linear, and that this is why the stabilization
is needed? Why is stabilization needed only for the SIA? I think these things should be made clear.}
\item \textcolor{blue}{Section 2.2.2: there is some imprecision here in equation (13) - you have not defined anywhere that $\Gamma$
is the boundary, and which boundary you are referring to. One can figure it out, but it is not precise.
$\Omega$ is not defined either though one will assume invariably that this is an open bounded domain in 2D
or 3D depending on which approximation one is looking at.}

These were not stated explicitly anywhere.
We've added them to table 1.
\item \textcolor{blue}{The boundary conditions are not discussed very rigorously systematically - the authors seem to
sprinkle in some boundary conditions here and there. I think the boundary conditions need to be
given for each of the models at the time the models are presented - boundary conditions are needed
to complete the definition of each models.}

We added this statement to section 2.1: ``We implement two types of boundary conditions for the prognostic equation.
Users can specify an inflow flux value and this value becomes a source of thickness at any point along the domain boundary where the ice velocity is pointing in to the domain.
The flux at the inflow boundary can change in time.
Second, we impose outflow boundary conditions on any part of the domain where the ice velocity is pointing outwards.
Which segments of the boundary are inflow or outflow are diagnosed automatically by calculating the sign of the dot product between the velocity and the unit outward normal vector.''

We also added an entirely new section which is now \S2.3 in the text just on the boundary conditions for the different diagnostic models.

\item \textcolor{blue}{Certain terms in the equations I do not believe are defined anywhere, for instance, in equation (10),
there is no expression given for the strains $\dot\epsilon(u)$.
This is one of the things I cam across that need to
be made more precise.}

@@

This was stated in table 1 but we've added the definition as $\frac{1}{2}(\nabla u + \nabla u^\top)$ and added a sentence to the text referring readers to table 1.
\item \textcolor{blue}{Section 2.2.3: the authors comment that higher degree polynomials can be used in the vertical layer
in the hybrid approach. What order is typically used?}

Added the following text: ``Going up to a degree-4 model is sufficient to capture the exact solution for the shallow ice approximation.
In the tutorial notebooks for icepack, we use up to degrees 2 and 4, but the test suite checks up to degree 8.''
\item \textcolor{blue}{Section 2.2.3: this might be a naive question, but does Glen's flow law come into the hybrid model?
I was expecting to see it there, but maybe I'm missing something.}
\item \textcolor{blue}{P. 10: the discussion here about substituting model components suggests it may be possible to use
different models in different regions and couple them (a la the ISCAL method). Is this possible, or
something that the authors are thinking to add to their model/code?}

Implementing this idea will require some new developments to Firedrake, namely first-class support for subdomains, defining different PDEs on different subdomains, and defining matching conditions for the solutions at the interfaces between subdomains.
The Firedrake developers are working on this feature right now as it's very much in demand.
We are very much interested in, for example, using SIA in the interior of the ice sheet and SSA in the ice streams and margins, as this would give a much less computationally-intensive way to do some form of whole-ice sheet modeling than using, say, the first order model.
\item \textcolor{blue}{Section 2.4: This section is very incomplete. You need to give the enthalpy equation and given the
Glen's law expression as well, since it is mentioned.}

We have added some text and equations describing the model we used, the boundary conditions, and the shear heating rate, the latter of which implicitly includes Glen's flow law.
\item \textcolor{blue}{In my opinion, there is not enough discussion of the thickness equation (Section 2.1) and how it
is discretized. In typical ice sheet models, this equation is used to change the ice extent - one
meshes up a region of ``potential'' ice, and then uses the thickness to dynamically determine a mask
for ice-covered regions. Do you do something like this in your model? It should be discussed for
completeness. I think you maybe start to do this in Section 4.1, but it is very confusing and hard to
make the connection.}

We added this statement to section 2.1: ``Icepack represents the thickness using continuous, piecewise polynomial basis functions in each cell of the mesh.
In the examples we use up to degree 2 and the unit tests use up to degree 4.
We have not yet implemented a formulation that works with discontinuous basis functions, but this extension is completely feasible within our framework.''
See also previous comment on boundary conditions for the thickness equation.
We also added a longer description of our treatment of ice-free regions (which is very ad hoc for now) at the end of section 2.1.
This is a weak point at present and we plan to improve this in future versions.

\item \textcolor{blue}{What sort of meshes do you use in your model (in the horizontal dimension, for the hybrid one)?
Structured/unstructured? Hex/tet (quad/tri)?}

@@

We use unstructured triangular meshes although Firedrake in principle can use unstructured quad meshes.
\item \textcolor{blue}{Section 3: It is not clear from the description what the inverse problem you are describing is for.
Is it to obtain parameters in the model like the basal friction using observational data of e.g. surface velocity? There really needs to be more discussion here, and I personally would like to see
a mathematical statement of a representative inverse problem you are solving. It would be worth
citing the work Perego et al. on optimization-based inversion, if what you are doing is similar:
https://doi.org/10.1002/2014JF003181. BTW, the basal friction has not been defined, yet it is
discussed - it needs to be defined earlier, when talking about boundary conditions (which needs to
be added).}

We added a brief description of the mathematics of the inverse problem to be solved.
We also added: ``The state to be estimated can be any single input field to the diagnostic model -- basal friction, rheology, or another field that the user has added by customizing the model.''

\item \textcolor{blue}{Section 4.1: I find this section confusing. I assume you are talking about discretizing the thickness
equation here - that should be made clear. I disagree with several statements in this section as
well. ``The simplest explicit timestepping schemes are unstable with CG finite elements'' - if you are
talking about CFL stability, this is not true. You need to satisfy a CFL condition which could give
rise to very small time-steps but you can get the scheme to be stable. I'm also confused about the
notion of SUPG as a time-stepping scheme - I think of SUPG as a finite element approach to deal
with advection-dominated flow problems, for example, that does not have anything to do with time-
stepping. Maybe you are referring to upwinding? In any case, I think SUPG has nothing to do with
forward Euler, so the discussion about forward Euler requiring parameters is erroneous. Additionally,
I don't understand the comment about implicit Euler smoothing out sharp discontinuities... I believe
explicit and implicit Euler have effectively the same diffusion and dispersion properties, so there
should not really be a difference between the schemes. Did the authors verify their time-stepper on
a manufactured problem to ensure that it was implemented correctly?}

This was sloppily written.
What we should have said was that SUPG confers some of the benefits of upwind finite difference stencils when using continuous Galerkin basis functions.
I was basing the statement about stability on the expression for the numerical amplification factor for the $\theta$-scheme with piecewise linear finite elements from section 3.5.2 in Donea and Huerta, Finite Element Methods for Flow Problems.
In any case, we've cut much of this section to focus more on the implementation in icepack rather than what other packages use.
The text was also out of date; the actual default now is an implicit scheme with a Lax-Wendroff correction that gives higher order accuracy in time.
Our statement that implicit Euler has predominantly diffusive errors was not to imply that explicit Euler doesn't share the same property.
We were trying to draw a contrast between what types of errors are tolerable for the prognostic model as opposed to other problems like damage transport, which is described in the next paragraph in the text.

\item \textcolor{blue}{Section 4.2: there is an approach discussed in Tezaur et al. (https://doi.org/10.5194/gmd-8-1197-2015)
for dealing with bad initial conditions in a Newton solver that relies on homotopy continuation that
would be worth citing. It is an alternate to the approach you describe that lets you get away with not
doing a line search for Newton. By the way, it should be no surprise that Newton is not converging
without a line search - in general Newton is not guaranteed to converge from an arbitrary initial
guess without the line search.}

@@

Added a reference to the Albany paper as well as another one on trust region methods.
We described a fairly rudimentary line search method in more detail than perhaps is necessary for a reader who's a seasoned modeler.
It's a bit of a pet peeve of mine when papers just say ``We used Newton!'' without any attention to the globalization strategy, which can make a huge difference to the solver robustness.
\item \textcolor{blue}{Section 4.3, lines 411-412: there are actually ways to construct weighted norms to deal with the issue
of DOFs having different orders of magnitude for the purpose of convergence.}

I think you can use a lumped mass matrix as the $H_0$ in BFGS too, but I find it to be far preferable to use something like Gauss-Newton which gives mesh-independent convergence and which achieves close to the second-order rate of full Newton on many problems.
No change to the text.
\item \textcolor{blue}{Section 4.4: Again, it is not clear to me what is your inverse problem. You need to state this explicitly
so it is clear.}

See correction to section 3.
The point of the inverse solver class is that it is very general with respect to what field is being inferred and what diagnostic model is being used.
Since the solver was designed to work for many different inverse problems
\item \textcolor{blue}{Section 4.5.1: you talk about problems defined on ``extruded geometries'' - do you ever use non-
extruded geometries? It has been shown in various references that there can be numerical problems
for land-ice solvers that do not use extruded geometries, e.g. Tezaur et al.}

We have restricted our implementation to use only extruded geometries.
Having made this choice, we might not ever be able to solve really geometrically complex problems like what the Elmer/Ice crowd did with the drainage of the lake underneath T\^ete Rousse glacier.
But the simplifications that this results in for the vast majority of glaciological applications are so advantageous that it's a sacrifice we're willing to accept.
\item \textcolor{blue}{Section 4.5.2: I think this section needs to be made earlier, and other BCs need to be added to that
discussion.}

See response to previous comment and the additional section we added on boundary conditions.
This section describes some extra care that we had to do in our implementation which was purely a consequence of our choice of basis functions and which does not appear in the idealized mathematical form of the model.
In keeping with our overall goal of splitting the paper up into a section on what we're solving and a different section on how we're solving it, we've kept this section where it is.

\item \textcolor{blue}{Section 4.6: The authors mention running their code on 1 core. Is the code parallel - can it be run
on multiple cores? Are there any hope for performance portability of the code to take advantage of
emerging HPC architectures, e.g., GPUs?}

### 2021-04-03 icepack.tex
More responses and describe BCs

The rheology and friction coefficient are functions of other fields that have their own evolution equations.
For example, the rheology is a function of temperature, englacial water content, and damage from crevassing.
Likewise, the friction coefficient can be described in terms of the subglacial water pressure and a roughness factor for the underlying bedrock \citep{cuffey2010physics}.
The diagnostic and prognostic equations can then be supplemented with evolution equations for these fields, for example the heat equation for temperature, or a hydrology model for subglacial water pressure.

@@

The \emph{prognostic model} or \emph{mass transport equation} (the two terms are synonymous) describes how the ice thickness changes in time.
The prognostic model is:
\begin{equation}
    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
\end{equation}
where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
This problem has the apparent form of a conservative advection equation.
The velocity $u$ is coupled to the thickness and surface slope, however, in such a way that the whole problem is not hyperbolic and in many cases does not have a PDE ``type'' in the usual sense.

Icepack represents the thickness using continuous, piecewise polynomial basis functions in each cell of the mesh.
We have not yet implemented a formulation that works with discontinuous basis functions, but this extension is completely feasible within our framework.

The surface elevation is calculated as
\begin{equation}
    s = \max\{b + h, (1 - \rho_I / \rho_W)h\},
\end{equation}
the first case corresponding to grounded ice and the second case corresponding to floating ice.
In the immediate vicinity of the grounding line, the assumption that the floating ice is in hydrostatic equilibrium with the ocean fails.
Most models assume hydrostasy and icepack does as well.
Elmer/Ice, on the other hand, solves a contact problem for the moving upper and lower ice surfaces and thus can accurately model non-hydrostatic ice shelves \citep{gagliardini2013capabilities}.
We have not implemented this feature but support for non-hydrostatic ice shelves may come in a future release of icepack.

@@

The most complex and most physically accurate model that we implement is the \emph{first-order} or \emph{Blatter-Pattyn} approximation, a 3D system describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
(In icepack we refer to our implementation of this equation as a \emph{hybrid model} as it includes both shear and plug flow modes, described below.)
The first-order model is based on an asymptotic expansion of the Stokes equations in the ratio of the ice thickness to a typical horizontal length scale.
The aspect ratio of most glacier flows is on the order of 1/20 or less, although there are some exceptions.
For example, the main trunk of Jakobshavn Isbrae in Greenland flows through a very deep and narrow trough with an aspect ratio closer to 1/5.

@@

\subsubsection{Shallow ice approximation}

@@

The shallow stream model can be derived from the Blatter-Pattyn approximation by assuming nearly plug flow, i.e. $\partial u/\partial z \ll \partial u/\partial x$.
The momentum equations can, again, be vertically integrated to obtain a 2D system of equations.
The class in icepack that represents this physics model is called \texttt{IceStream}.
The terms in the action functional are:
\begin{align}
    \text{viscosity} & = \frac{n}{n + 1}\int_\Omega hA^{-\frac{1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}dx \\
    \text{friction} & = \frac{m}{m + 1}\int_\Omega C|u|^{\frac{1}{m} + 1}dx \\
    \text{gravity} & = -\int_\Omega\rho_I gh\nabla s\cdot u\,dx \\
    \text{terminus} & = \frac{1}{2}\int_\Gamma(\rho_I gh^2 - \rho_Wgd^2)u\cdot \nu\,d\gamma
\end{align}
and the action functional is
\begin{equation}
    J = \text{viscosity} + \text{friction} - \text{gravity} - \text{terminus}.
\end{equation}
When the ice is floating, there are two simplifications: the friction coefficient $C$ is 0, and the surface elevation $s$ can be written in terms of the thickness $h$ as
\begin{equation}
    s = (1 - \rho_I / \rho_W)h.
\end{equation}
Additionally, the terminal stress term of the action disappears after applying integration by parts to the gravity term to shift the gradient of the surface elevation over onto the velocity.
Since the action functional for ice shelves has fewer terms than for grounded ice streams, we have defined a separate \texttt{IceShelf} model class.
The ice shelf and ice stream models share common components, i.e. the viscosity and side wall stress.

@@

@@

We implemented the enthalpy transport model described in \citet{aschwanden2012enthalpy}.
The enthalpy density $E$ describes the heat content of the material in a way that incorporates both temperature $T$ and latent heat stored in meltwater:
\begin{equation}
    E = \rho_I(cT + Lf)
\end{equation}
where $L$ is the latent heat of melting of ice and $f$ is the melt fraction.
The melt fraction can only be positive when the ice as at the pressure-melting point.
The temperature and meltwater fraction can be uniquely calculated at any point from the value of the enthalpy, so nothing is sacrificed in describing heat content one way or the other.
Using the enthalpy has the advantage of circumventing many of the difficulties associated with tracking the interface between cold and temperate ice.
Our implementation of the model uses all of the simplifying assumptions described in \citet{aschwanden2012enthalpy}, for example that horizontal diffusion is negligible and that heat capacity and conductivity are not temperature-dependent within each phase.
The resulting PDE is
\begin{equation}
    \left(\frac{\partial}{\partial t} + \nabla\cdot u\right)E - \frac{\partial}{\partial z}\alpha\frac{\partial E}{\partial z} = q
\end{equation}
where $u$ is the full 3D velocity, $\alpha$ the (temperature-dependent) thermal diffusivity, and $q$ the volumetric heat sources.
This form assumes some amount of vertical diffusion of water at the melting point.
A realistic treatment would require treating the ice as a porous medium and modeling englacial water transport explicitly, which we have not done.
The boundary condition at the ice base is a fixed flux of heat from geothermal sources or from the oceans.
We depart from \citet{aschwandend2012enthalpy} in one respect.
Rather than fix the ice surface temperature to the atmospheric temperature, we instead use a Robin boundary condition that makes the ice temperature adjust to the external temperature:
\begin{equation}
    -\alpha\frac{\partial E}{\partial z}\Big|_{z = s} = \kappa(E - E_s)
\end{equation}
where $\kappa$ is a surface exchange coefficient and $E_s$ is the temperature-equivalent enthalpy that forces the surface.
Using a Robin instead of a Dirichlet boundary condition at the surface allows for more gradual adjustment to above-freezing temperatures that might otherwise create areas of unrealistically high melt fraction.
Except in blue ice areas, the thermal contact between ice and the atmosphere is mediated by firn, which we do not model explicitly.
The Robin boundary condition is a compromise to work around this missing component of the system.
Finally, if there is an upstream boundary the user must provide the inflow values of the enthalpy with depth, while outflow boundary conditions are applied at the glacier terminus.

@@

The simplest explicit timestepping schemes are unstable with continuous Galerkin finite elements.
To solve the advection equation, other packages use the streamlined upwind Petrov-Galerkin (SUPG) method for the timestepping scheme \citep{brinkerhoff2013data, larour2012continental}.
The SUPG method with an explicit time discretization is conditionally stable \citep{donea2003finite}, but this scheme introduces a tuneable stabilization parameter.
In the interest of simplicity, we instead default to the implicit Euler scheme, which has no such parameters.
The implicit Euler scheme requires solving a non-symmetric linear system, but the computational cost comes with the advantage of unconditional stability.

Some glaciologists may be unfamiliar with the Courant-Friedrichs-Lewy (CFL) condition.
The use of an unconditionally stable scheme guarantees that they will get an answer, rather than a runtime error, should they try to use a large timestep.
The extra computational cost of using an implicit time discretization for the prognostic equation is dwarfed by the cost of the diagnostic solve in any case.
Advanced users who are interested in maximizing performance can subclass the solver to implement a faster explicit scheme.

Icepack currently lacks an adaptive timestepping scheme.
Unconditionally stable schemes allow taking long timesteps, but taking very long timesteps will give inaccurate solutions.
At present, users are still responsible for checking the accuracy of their results, for example by running at more than one resolution.
Adaptive timestepping will be added in a future release.

The implicit Euler scheme tends to diffuse out sharp discontinuities that may be present in the true solution \citep{donea2003finite}.
Since the ice thickness does not possess shockwaves or propagating discontinuities this error mode is tolerable.
The coupling of ice thickness to velocity makes the whole system more resemble a parabolic problem than a hyperbolic one, and under the shallow ice approximation the system is truly parabolic.

Other problems in glaciology have more of a hyperbolic character.
For example, the thresholding behavior of the source terms for the damage model (see \S\ref{sec:damage-transport}) can create sharp discontinuities.
The implicit Euler scheme would obscure this important feature.
For the damage solver, we have instead used a strong stability-preserving Runge-Kutta method in time and a discontinuous Galerkin basis in space to best capture these sharp interfaces \citep{shu1988efficient}.

@@

\subsubsection{Ocean boundary condition}

@@

@@

We have also paid special attention to how we can design this software package to be most usable for its intended audience.
Relatively few works in the computational science literature draw directly from relevant work in HCI when discussing usability; see for example \citet{hannay2009scientists, harris2020array}.
We believe that this is because of two difficulties: (1) evaluating the degree to which usability is a rate-limiting factor and (2) concretely assessing what features make software tools more or less usable.
By contrast, measuring computational performance is much more feasible although still fraught with difficulties of its own.
(This is not to say that performance optimization is easy by any means.)
In working with several graduate students and postdoctoral researchers in glaciology, we have observed that usability is a substantial bottleneck for researchers at these career stages.
For this reason, we have chosen to focus explicitly on usability by drawing on the research literature in HCI.
We argue that the same design features that enhance usability for relative novices to the subject area will also enhance the productivity of expert users.

This paper has presented several physics models currently implemented in icepack along with demonstrations.
Future developments will include:
\begin{enumerate}
    \item an implementation of full Stokes flow,
    \item dynamic glacier termini to enable calving models, and
    \item adaptivity through moving mesh methods.
\end{enumerate}
New features will be guided by feedback from icepack users and from the glaciological community at large.
By providing implementations of several glacier flow models from less to more complex and by enabling users to experiment with the physics, this tool both lowers the barrier to entry for novices to glacier flow modeling and provides a pathway for these users to progress towards ever more sophisticated and advanced simulations.

>>>>

The evolution equations determining for ice thickness and velocity are necessary for any realistic glacier simulation, but other fields with their own evolution equations can be a part of the problem as well.
The rheology and friction are the main coefficients in the diagnostic equation.
These coefficients are functions of other fields that have their own evolution equations.
For example, the rheology is a function of temperature, englacial water content, and damage from crevassing.
Likewise, the friction coefficient can be described in terms of the subglacial water pressure and a roughness factor for the underlying bedrock \citep{cuffey2010physics}.
The diagnostic and prognostic equations can then be supplemented with evolution equations for these fields, for example the heat equation for temperature, or a hydrology model for subglacial water pressure.

@@

The \emph{prognostic model} or \emph{mass transport equation} (the two terms are synonymous) describes how the ice thickness changes in time.
The prognostic model is:
\begin{equation}
    \frac{\partial h}{\partial t} + \nabla\cdot hu = \dot a_s - \dot a_b,
    \label{eq:prognostic-eqn}
\end{equation}
where $\dot a_s$ and $\dot a_b$ are the surface and basal mass balance.
This problem has the apparent form of a conservative advection equation, but the velocity $u$ is coupled to the thickness and surface slope in such a way that the whole problem is not hyperbolic.
For the specific case of the shallow ice approximation (see section \S\ref{subsubsec:shallow-ice}), the coupled system is parabolic.
In all other cases, the problem does not have a PDE ``type'' in the usual sense because the velocity is found through solving an elliptic PDE where the thickness and surface slope are coefficients.

Icepack represents the thickness using continuous, piecewise polynomial basis functions in each cell of the mesh.
In the examples we use up to degree 2 and the unit tests use up to degree 4.
We have not yet implemented a formulation that works with discontinuous basis functions, but this extension is completely feasible within our framework.

The surface elevation is calculated as
\begin{equation}
    s = \max\{b + h, (1 - \rho_I / \rho_W)h\},
\end{equation}
the first case corresponding to grounded ice and the second case corresponding to floating ice.
In the immediate vicinity of the grounding line, the assumption that the floating ice is in hydrostatic equilibrium with the ocean fails.
Most models assume hydrostasy and icepack does as well.
Elmer/Ice, on the other hand, solves a contact problem for the moving upper and lower ice surfaces and thus can accurately model non-hydrostatic ice shelves \citep{gagliardini2013capabilities}.

We implement two types of boundary conditions for the prognostic equation.
Users can specify an inflow flux value and this value becomes a source of thickness at any point along the domain boundary where the ice velocity is pointing in to the domain.
The flux at the inflow boundary can change in time.
Second, we impose outflow boundary conditions on any part of the domain where the ice velocity is pointing outwards.
Which segments of the boundary are inflow or outflow are diagnosed automatically by calculating the sign of the dot product between the velocity and the unit outward normal vector.

The mass transport equation for ice thickness is a free boundary problem, where the free boundary is the contour between ice-covered and ice-free regions \citep{schoof2013ice}.
A naively-implemented prognostic solver could erroneously compute negative thickness values in subsequent timesteps when there is ablation in ice-free regions.
A common and ad-hoc approach to work around this issue is to truncate the thickness at zero at every timestep.
The principled approach is to instead treat the free boundary problem directly as a variational inequality \citep{jouvet2012steady}.
Icepack currently lacks a principled scheme for tracking this free boundary and we instead rely on truncation.
Treating this problem as a variational inequality in icepack will be the subject of future development.
PETSc includes scalable solvers for variational inequalities \citep{bueler2020petsc} that are also available through Firedrake.

@@

The most complex and most physically accurate model that we implement is the \emph{first-order} or \emph{Blatter-Pattyn} approximation, a 3D system describing the horizontal velocity \citep{blatter1995velocity, pattyn2003new}.
(In icepack we refer to our implementation of this equation as a \emph{hybrid model} as it includes both shear and plug flow modes, described below.)
The first-order model is based on an asymptotic expansion of the Stokes equations with respect to the ratio of the ice thickness to a typical horizontal length scale.
The aspect ratio of most glacier flows is on the order of 1/20 or less, although there are some exceptions.
For example, the main trunk of Jakobshavn Isbrae in Greenland flows through a very deep and narrow trough with an aspect ratio closer to 1/5.

@@

\subsubsection{Shallow ice approximation}\label{subsubsec:shallow-ice}

@@

The shallow stream model can be derived from the Blatter-Pattyn approximation by assuming nearly plug flow, i.e. $\partial u/\partial z \ll \partial u/\partial x$.
The momentum equations can, again, be vertically integrated to obtain a 2D system of equations.
The class in icepack that represents this physics model is called \texttt{IceStream}.
The terms in the action functional are:
\begin{align}
    \text{viscosity} & = \frac{n}{n + 1}\int_\Omega hA^{-\frac{1}{n}}|\dot\varepsilon(u)|^{\frac{1}{n} + 1}dx \\
    \text{friction} & = \frac{m}{m + 1}\int_\Omega C|u|^{\frac{1}{m} + 1}dx \\
    \text{gravity} & = -\int_\Omega\rho_I gh\nabla s\cdot u\,dx \\
    \text{terminus} & = \frac{1}{2}\int_\Gamma(\rho_I gh^2 - \rho_Wgd^2)u\cdot \nu\,d\gamma \label{eq:terminus-bc}
\end{align}
and the action functional is
\begin{equation}
    J = \text{viscosity} + \text{friction} - \text{gravity} - \text{terminus}.
\end{equation}
When the ice is floating, there are two simplifications: the friction coefficient $C$ is 0, and the surface elevation $s$ can be written in terms of the thickness $h$ as
\begin{equation}
    s = (1 - \rho_I / \rho_W)h.
\end{equation}
Additionally, the terminal stress term of the action disappears after applying integration by parts to the gravity term to shift the gradient of the surface elevation over onto the velocity.
Since the action functional for ice shelves has fewer terms than for grounded ice streams, we have defined a separate \texttt{IceShelf} model class.
The ice shelf and ice stream models share common components, i.e. the viscosity and side wall stress.

@@

\subsection{Boundary conditions}

The shallow shelf, shallow stream, and hybrid models require boundary conditions to be well-posed.
We implement three different types which are largely the same for each physics model.
\begin{enumerate}
    \item Users can specify the velocity on segments of the domain boundary where ice is flowing in; this is a Dirichlet condition which is eliminated from the system at the level of the solver.
Supplying inflow boundary conditions for the hybrid model necessarily requires more information, namely the variation of the horizontal velocity with depth.
    \item The Neumann condition at the ice terminus (equation \eqref{eq:terminus-bc}) is a natural boundary condition which is imposed by adding the relevant term to the action functional.
    \item We allow for side wall boundary conditions where the ice has non-zero thickness but abuts, say, a fjord wall or an ice shelf embayment that exerts resistive stresses.
In the normal direction, the boundary condition is that $u \cdot \nu = 0$ where $\nu$ is the unit outward normal vector.
In the tangential direction, the boundary condition is $hM\cdot\nu = -C|u|^{\frac{1}{m} - 1}u$ where $C$ is a side wall friction coefficient and $M$ is the membrane stress tensor.
Putting these together gives a mixed Dirichlet-Robin boundary condition.
The side wall friction (Robin) boundary condition is natural and thus can be added directly to the action functional.
The constraint of no normal flow is harder to impose directly for a curved boundary; we used the penalty method.
\end{enumerate}
Future versions of the code will use Nitsche's method \citep{nitsche1971variationsprinzip} instead of the penalty method to enforce the no normal flow constraint.
Nitsche's method dramatically improves the conditioning of the associated optimization problem.
Our implementation of the hybrid model requires special care to handle the terminus boundary condition as a consequence of our choice of vertical basis functions; see section \S\ref{subsubsec:ocean-bc}.

The shallow ice model alone is able to cope with ice-free regions in the domain, albeit with truncation at 0 thickness.
All of the remaining models and their boundary conditions assume that the entire spatial domain is ice-covered.
We will improve the solvers to address ice-free areas in future versions.

@@

We implemented the enthalpy transport model described in \citet{aschwanden2012enthalpy}.
The enthalpy density $E$ describes the heat content of the material in a way that incorporates both temperature $T$ and latent heat stored in meltwater:
\begin{equation}
    E = \rho_I(cT + Lf)
\end{equation}
where $L$ is the latent heat of melting of ice and $f$ is the melt fraction.
The melt fraction can only be positive when the ice as at the pressure-melting point.
The temperature and meltwater fraction can be uniquely calculated at any point from the value of the enthalpy, so nothing is sacrificed in describing heat content one way or the other.
Using the enthalpy has the advantage of circumventing many of the difficulties associated with tracking the interface between cold and temperate ice.
Our implementation of the model uses all of the simplifying assumptions described in \citet{aschwanden2012enthalpy}, for example that horizontal diffusion is negligible and that heat capacity and conductivity are not temperature-dependent within each phase.
The resulting PDE is
\begin{equation}
    \left(\frac{\partial}{\partial t} + \nabla\cdot u\right)E - \frac{\partial}{\partial z}\alpha\frac{\partial E}{\partial z} = q
\end{equation}
where $u$ is the full 3D velocity, $\alpha$ the (temperature-dependent) thermal diffusivity, and $q$ the volumetric heat sources.
This form assumes some amount of vertical diffusion of water at the melting point.
A realistic treatment would require treating the ice as a porous medium and modeling englacial water transport explicitly, which we have not done.
The boundary condition at the ice base is a fixed flux of heat from geothermal sources or from the oceans.
We depart from \citet{aschwanden2012enthalpy} in one respect.
Rather than fix the ice surface temperature to the atmospheric temperature, we instead use a Robin boundary condition that makes the ice temperature adjust to the external temperature:
\begin{equation}
    -\alpha\frac{\partial E}{\partial z}\Big|_{z = s} = \kappa(E - E_s)
\end{equation}
where $\kappa$ is a surface exchange coefficient and $E_s$ is the temperature-equivalent enthalpy that forces the surface.
Using a Robin instead of a Dirichlet boundary condition at the surface allows for more gradual adjustment to above-freezing temperatures that might otherwise create areas of unrealistically high melt fraction.
Except in blue ice areas, the thermal contact between ice and the atmosphere is mediated by firn, which we do not model explicitly.
The Robin boundary condition is a compromise to work around this missing component of the system.
Finally, if there is an upstream boundary the user must provide the inflow values of the enthalpy with depth, while outflow boundary conditions are applied at the glacier terminus.

@@

Icepack offers two timestepping schemes for solving the prognostic model (equation \eqref{eq:prognostic-eqn}).
For testing purposes we include the implicit Euler scheme, which is first-order accurate and unconditionally stable for the advection equation.
The default is a second-order accurate scheme based on an implicit version of the Lax-Wendroff method, which is also unconditionally stable for the advection equation \citep{donea2003finite}.
Explicit schemes all require a timestep that satisfies the Courant-Friedrichs-Lewy (CFL) stability condition, which many glaciologists may be unfamiliar with.
With an unconditionally stable scheme, users will get an answer, rather than a runtime error, should they try to use a larger timestep.
The extra computational cost of using an implicit time discretization for the prognostic equation is dwarfed by the cost of the diagnostic solve.
Advanced users who are interested in maximizing performance can subclass the solver to implement a faster explicit scheme.
We note that, while we the stability properties of different schemes for the linear advection equation guided our choice of method, the coupled system for both thickness and velocity is not linear and not hyperbolic.

The implicit Euler and Lax-Wendroff schemes tend to diffuse out sharp discontinuities that may be present in the true solution \citep{donea2003finite}.
Since the ice thickness does not possess shockwaves or propagating discontinuities this error mode is tolerable.
The coupling of ice thickness to velocity makes the whole system more resemble a parabolic problem than a hyperbolic one, and under the shallow ice approximation the system is truly parabolic.

Other problems in glaciology have more of a hyperbolic character.
For example, the thresholding behavior of the source terms for the damage model (see \S\ref{sec:damage-transport}) can create sharp discontinuities.
The implicit Euler scheme would obscure this important feature.
For the damage solver, we have instead used a strong stability-preserving Runge-Kutta method in time, a discontinuous Galerkin basis in space, and a flux limiter to best capture these sharp interfaces \citep{shu1988efficient}.

Icepack currently lacks an adaptive timestepping scheme.
Implicit schemes allow taking longer timesteps than explicit ones, but taking very long timesteps will give inaccurate solutions and, in the presence of ablation, may yield negative thickness values.
At present, users are still responsible for checking the accuracy of their results, for example by running at more than one resolution.
Adaptive timestepping will be added in a future release.

@@

\subsubsection{Ocean boundary condition}\label{subsubsec:ocean-bc}

@@

Simulations that have been debugged can then easily be transformed into a single Python script, for example using the tool \texttt{nbconvert}, for production runs and parallel execution.
In other words, while there is an interactive interface, there is also a faster batch mode interface as well.

@@

We have also paid special attention to how we can design this software package to be most usable for its intended audience.
Relatively few works in the computational science literature draw directly from relevant work in HCI when discussing usability; see for example \citet{hannay2009scientists, harris2020array}.
We believe that this is because of two difficulties.
First, the degree to which usability is a rate-limiting factor for scientists is hard to quantify and likely differs widely across disciplines.
Second, concretely assessing what features make software tools more or less usable is highly subjective.
By contrast, measuring computational performance is much more feasible although still fraught with difficulties of its own.
(This is not to say that performance optimization is easy by any means.)
In working with several graduate students and postdoctoral researchers in glaciology, we have observed that usability is a substantial bottleneck for researchers at these career stages.
For this reason, we have chosen to focus explicitly on usability by drawing on the research literature in HCI.
We argue that the same design features that enhance usability for relative novices to the subject area will also enhance the productivity of expert users.

This paper has presented several physics models currently implemented in icepack along with demonstrations.
Future developments will include:
\begin{enumerate}
    \item an implementation of full Stokes flow,
    \item improved physics formulations and solvers that work in ice-free areas,
    \item adaptivity in time and space,
\end{enumerate}
New features will be guided by feedback from icepack users and from the glaciological community at large.
By providing implementations of several glacier flow models from less to more complex and by enabling users to experiment with the physics, this tool both lowers the barrier to entry for novices to glacier flow modeling and provides a pathway for these users to progress towards ever more sophisticated and advanced simulations.
